Efficient Machine Learning in a Memristor Network

### 2025-06-05 paper/paper.tex
Paper beofre incorporating comments

A modification to any of these weights affects the network's behaviour by altering the functional form of the memristor’s steady-state conductance, $g_{\infty}(\boldsymbol{\Delta})$, which depends on the local potential, pressure, and concentration drops across the device. Importantly, the \emph{entire functional shape} of $g_{\infty}$ plays a role in determining the network’s dynamics and ultimate steady state, as illustrated in \cref{app:dyn_evol_steady_state}.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=1.7\columnwidth]{plots/voltage_divider/mse_weights_vd.pdf}
    \caption{(a) Behaviour of the cost function during the training of the memristor voltage divider. Different colors indicate that the training is done with a specific choice of training weights, as indicated in the legend. Learning rates are set to $[3\cdot 10^{-6}, 3\cdot 10^{-6}, 5\cdot 10^{-4}, 100]$. (b) Pressures imposed on the three nodes of the voltage divider during training. The inset shows the relaxation to the steady state of the output voltage $V_2$ to the desired value $V_D$ when the pressures imposed are the values obtained in the last training step, and the input nodes are stimulated with fixed voltage. (c) Ion concentration densities imposed on the three nodes of the voltage divider during training. (d) Length of the memristors during training. (e) Base radius of the memristors during training.}\label{fig:mse_weights_vd}
\end{figure*} 
In the forward phase, the network is fed with specific electric potentials in the input nodes. Following both circuit laws and memristor's dynamics, the network reaches a steady state where electrical potentials can be measured in the output nodes. Clearly, the output of the network depends on the set of weights imposed on the network $\boldsymbol{w}$, regardless them being node or edge defined. The value of the outputs can be used to quantify the status of the network, meaning how much its answer differs from the desired one. Specifically, we define a cost function that computes the difference squared between the voltage in the output nodes and the values of desired voltages
\[
C\left( \boldsymbol{w} \right) = \sum_{i \in \Lambda} \brac{V_i\brac{\boldsymbol{w}} - V_i^D}^2,
\label{eq:cost_func}
\]
where $\Lambda$ represents the set of indices belonging to the output nodes, $V_i\brac{\boldsymbol{w}}$ the voltage measured on the output nodes and $V^D$ is a vector containing the values of desired output nodes.

@@

\subsection{\label{sec:train_memr_vd}Training a Memristor Voltage Divider}
A voltage divider is an electrical circuit made of three nodes connected in series by two resistances, that returns a fraction of the input voltage. Given the simplicity of its structure, we use it as an initial playground for the training algorithm described above. Specifically, we replace the resistances with memristor channels, to be trained such that giving $V_1 = 5V$ and $V_3 = 0V$ as inputs, we get $V_2 = 4V$ as output. In \cref{fig:vd_scheme} a schematic of the memristor voltage divider is shown.
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\columnwidth]{plots/voltage_divider/vd_scheme.pdf}
    \caption{Schematical representation of a memristor voltage divider.}\label{fig:vd_scheme}
\end{figure}

We initialize the system with pressures $P_1 =  P_2 = 1 \text{bar}$ and ion concentration $\rho_1 = \rho_2 = 0.2 \text{mM}$ on each node, length $L_1 = L_2 = 10\mu \text{m}$ and base radius  $R_{b1} = R_{b2} = 200 \text{nm}$ on each edge. After choosing one of these four physical quantities as training parameter $\boldsymbol{w} = \brac{w_1, w_2}$, we implement the training as described in \cref{sec:training} that modifies them, while the other parameters are kept constant. 
\begin{figure}[b]
    \centering
    \includegraphics[width=0.8\columnwidth]{plots/voltage_divider/evolution_targets.pdf}
    \caption{(a) The training of a memristor voltage divider to a set of desired voltage outputs, indicated by dashed light purple lines. The output voltage $V_2$, indicated by dark purple dots, adapts to the desired output during the training. The input voltages $V_1$ and $V_3$ are constant at $5V$ and $0V$ respectively. (b) The dots with different shades of grey show the values of pressures, chosen as weights, imposed on the three nodes during training.}\label{fig:evolution_targets}
\end{figure}
In \cref{fig:mse_weights_vd}(a) we show that the cost function normalized to its initial value decreases of many order of magnitude during the training for each of the chosen weight. The values of the chosen set of weights simultaneously changes from the initial condition, as shown in \cref{fig:mse_weights_vd}(b)-(e), adjusting the behaviour of the network to the desired one. If the values of a chosen type of weight at the end of the training are imposed to the network together with the voltage inputs, the network will autonomously relax to a steady state that coincides with the desired state. To exemplify this point, the inset in \cref{fig:mse_weights_vd}(b) shows the physical relaxation of the voltage divider when the trained values of pressure, which correspond to the three values in the last training step, are imposed on the nodes together with the constant inputs voltages. The inset shows that the output voltage $V_2$ autonomously relaxes to the desired value of $V_D=4V$ in a time scale characteristic of the memristor.

In order to investigate the adaptability of the algorithm, we trained the system with a set of three consequent desired voltages $V_D=\{1V, 3V, 4V\}$. In \cref{fig:evolution_targets} we show how the output voltage $V_2$ subsequently adapts to the desired voltages during the training of the pressures in the nodes, the evolution of which can be seen in \cref{fig:evolution_targets}(b). It is evident that the system can be retrained at any stage to a new desired objective making it reusable. Any other weight type can be chosen, depending on the necessity.

@@

% \cleardoublepage
\subsection{\label{sec:linear_regression}Linear Regression}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\textwidth]{plots/regression/regression_G00010001.pdf}
    \caption{}\label{fig:lin_regr1}
\end{figure*}

@@

The performance of the algorithm on this task is shown in \cref{fig:lin_regr2}(b), where the learning rates for the different weights have been fine-tuned to prevent overshooting while still maintaining comparable convergence speed. From the behaviour of the cost function, it is evident that the choice of weight type has a decisive impact on the network’s ability to learn a complex task. Specifically, training the lengths of the channels on the edges results in successful convergence, whereas training the node pressures proves to be extremely inefficient. For a visual interpretation of the regression cost function, \cref{fig:lin_regr1}(c) shows the state of the training at three different time steps.
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\textwidth]{plots/regression/regression_G00010002.pdf}
    \caption{}\label{fig:lin_regr2}
\end{figure*}

To study the influence of memristor orientation on training performance, we retained the same network geometry but inverted the orientation of some memristors, as shown in \cref{fig:lin_regr2}(a). Specifically, if in the original configuration the base of a memristor pointed to one node and the tip to another, we reversed this direction.

This modification significantly affects the behaviour of the network. As shown in \cref{fig:lin_regr2}(b), the new orientation alters the trainability of different weight types: in this configuration, training node concentration leads to better performance, while training edge length becomes less efficient. The underlying reason is that inverting the orientation of a channel also inverts the sign of the external stimulus applied to it, since the stimuli are defined node-wise and the difference is always taken as "base minus tip."

This reversal changes both the shape and the conductivity range over which the function $g_{\infty}(\boldsymbol{\Delta})$ is defined, thereby limiting or enabling the conductivity values the channel can attain. A network with well-oriented channels can therefore access conductivity ranges that make the training solvable, while a poorly oriented network may be constrained to regions where solutions are not found. To exemplify this point, Figure~(X) of Ref.~(Y) shows $g_{\infty}$ as a function of pressure for positive and negative potential drops, assuming a zero concentration drop. The plot reveals two completely disjoint conductivity ranges depending on the sign of the potential drop, which explains why training on pressure alone may be challenging. This highlights the need for deeper exploration of network geometry when selecting the type of weight to train. In this regard, we observe that the edges whose orientation was inverted are directly connected to either the input or the output nodes. This suggests that particular attention should be given to these edges in future explorations, as they may have a bigger impact on trainability.
As for the length-based training, \cref{fig:lin_regr2}(c) illustrates how the network responds to the test set at three different stages of the training process.

Bigger networks have more degrees of freedom, and should allow for a bigger range of exploration of a solution. However, the experimental realization would become more challenging. In this work, we show that an adequate geometry for a specific training goal can be sufficient to make a small network trainable. A deeper analysis on the influence of geometry on the trainability is needed in the research contest of physical learning and can be explored in future research.

Larger networks offer more degrees of freedom and can, in principle, enable broader exploration of the solution space. However, their experimental realization becomes increasingly complex. In this work, we showed that an appropriately chosen geometry adapted to a specific training objective can be sufficient to make even a small networks trainable. A deeper analysis of the role of geometry in determining trainability is still needed in the context of physical learning.

@@

We proposed a physical neural network composed of nodes connected by aqueous memristors. These devices exhibit conductance that responds to external stimuli of various physical natures, such as electrical, mechanical, and chemical. Additionally, their conductivity is sensitive to changes in the geometric parameters of their channel-like structure.

Our motivation for using these devices lies in the ability to exploit these dependencies to control the conductance of each memristor in the network. This enables us to manipulate an input electrical signal in a programmable way, according to a specific computational objective.

@@

Training larger and more complex memristor networks is also feasible with the algorithm we proposed. We demonstrated that a network composed of 9 nodes and 15 edges can be successfully trained for input-output voltage mapping using all the weight types considered.

We then challenged a more complex network, with 10 nodes and 20 edges, to perform a linear regression task with an offset. In this setting, we observed that the trainability of the network using a specific weight type strongly depends on the network's geometry. First, we showed that different geometries yield different final training accuracies when using the same weight type. Then, keeping the geometry fixed, we demonstrated that the orientation of the memristors also plays a crucial role: reversing the direction of just three memristors connected to either input or output nodes improved the training precision when using concentration as the trained weight by three orders of magnitude.

Given that the expressiveness of deep neural networks grows exponentially with depth\cite{big_neuralnet1, big_neuralnet2}, we interpret the limitations encountered in training linear regression tasks as a consequence of the small size of the network. However, small-scale networks are experimentally attractive, and in such cases, a training error of a few orders of magnitude may be acceptable. A key challenge for scaling such networks lies in the physical control of individual weights, especially when the weight type involves mechanical or geometric parameters. Emerging technologies in microfluidics, soft robotics, and programmable matter may provide practical tools to overcome these limitations and scale the network while preserving control.

>>>>

A modification to any of these weights affects the network's behaviour by altering the functional form of the memristor’s steady-state conductance, $g_{\infty}(\boldsymbol{\Delta})$, which depends on the local potential, pressure, and concentration drops across the device. Importantly, the functional shape of $g_{\infty}$ plays a role in determining the network’s dynamics and ultimate steady state, as illustrated in \cref{app:dyn_evol_steady_state}.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=1.7\columnwidth]{plots/voltage_divider/mse_weights_vd.pdf}
    \caption{(a) Behaviour of the cost function during the training of the memristor voltage divider. Different colors indicate that the training is done with a specific choice of training weights, as indicated in the legend. Learning rates are set to $[3\cdot 10^{-6}, 3\cdot 10^{-6}, 5\cdot 10^{-4}, 100]$. (b) Pressures imposed on the three nodes of the voltage divider during training. The inset shows the relaxation to the steady state of the output voltage $V_2$ to the desired value $V_2^D$ when the pressures imposed are the values obtained in the last training step, and the input nodes are stimulated with fixed voltage. (c) Ion concentration densities imposed on the three nodes of the voltage divider during training. (d) Length of the memristors during training. (e) Base radius of the memristors during training.}\label{fig:mse_weights_vd}
\end{figure*} 
In the forward phase, the network is fed with specific electric potentials in the input nodes. Following both circuit laws and memristor's dynamics, the network reaches a steady state where electrical potentials can be measured in the output nodes. Clearly, the output of the network depends on the set of weights imposed on the network $\boldsymbol{w}$, regardless them being node or edge defined. The value of the outputs can be used to quantify the status of the network, meaning how much its answer differs from the desired one. Specifically, we define a \emph{cost function} that computes the difference squared between the voltage in the output nodes and the values of desired voltages
\[
C\left( \boldsymbol{w} \right) = \sum_{i \in \Lambda} \brac{V_i\brac{\boldsymbol{w}} - V_i^D}^2,
\label{eq:cost_func}
\]
where $\Lambda$ represents the set of indices belonging to the output nodes, $V_i\brac{\boldsymbol{w}}$ the voltage measured on the output nodes and $V^D$ is a vector containing the values of desired output nodes.

@@

\subsection{\label{sec:train_memr_vd}Training a Memristor Voltage Divider}
A voltage divider is an electrical circuit made of three nodes connected in series by two resistances, that returns a fraction of the input voltage. Given the simplicity of its structure, we use it as an initial playground for the training algorithm described above. Specifically, we replace the resistances with memristor channels, to be trained such that giving $V_1 = 5V$ and $V_3 = 0V$ as inputs, we get $V_2^D = 4V$ as output. In \cref{fig:vd_scheme} a schematic of the memristor voltage divider is shown.
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\columnwidth]{plots/voltage_divider/vd_scheme.pdf}
    \caption{Schematical representation of a memristor voltage divider.}\label{fig:vd_scheme}
\end{figure}

We initialize the system with pressures $P_1 =  P_2 = 1 \text{bar}$ and ion concentration $\rho_1 = \rho_2 = 0.2 \text{mM}$ on each node, length $L_1 = L_2 = 10\mu \text{m}$ and base radius  $R_{b1} = R_{b2} = 200 \text{nm}$ on each edge. After choosing one of these four physical quantities as training parameter $\boldsymbol{w} = \brac{w_1, w_2}$, we implement the training as described in \cref{sec:training} that modifies them, while the other parameters are kept constant. 
\begin{figure}[b]
    \centering
    \includegraphics[width=0.8\columnwidth]{plots/voltage_divider/evolution_targets.pdf}
    \caption{(a) The training of a memristor voltage divider to a set of desired voltage outputs, indicated by dashed light purple lines. The output voltage $V_2$, indicated by dark purple dots, adapts to the desired output during the training. The input voltages $V_1$ and $V_3$ are constant at $5V$ and $0V$ respectively. (b) The dots with different shades of grey show the values of pressures, chosen as weights, imposed on the three nodes during training.}\label{fig:evolution_targets}
\end{figure}
In \cref{fig:mse_weights_vd}(a) we show that the cost function normalized to its initial value decreases of many order of magnitude during the training for each of the chosen weight. The values of the chosen set of weights simultaneously changes from the initial condition, as shown in \cref{fig:mse_weights_vd}(b)-(e), adjusting the behaviour of the network to the desired one. If the values of a chosen type of weight at the end of the training are imposed to the network together with the voltage inputs, the network will autonomously relax to a steady state that coincides with the desired state. To exemplify this point, the inset in \cref{fig:mse_weights_vd}(b) shows the physical relaxation of the voltage divider when the trained values of pressure, which correspond to the three values in the last training step, are imposed on the nodes together with the constant inputs voltages. The inset shows that the output voltage $V_2$ autonomously relaxes to the desired value of $V_2^D=4V$ in a time scale characteristic of the memristor.

In order to investigate the adaptability of the algorithm, we trained the system with a set of three consequent desired voltages $V_2^D=\{1V, 3V, 4V\}$. In \cref{fig:evolution_targets} we show how the output voltage $V_2$ subsequently adapts to the desired voltages during the training of the pressures in the nodes, the evolution of which can be seen in \cref{fig:evolution_targets}(b). It is evident that the system can be retrained at any stage to a new desired objective making it reusable. Any other weight type can be chosen, depending on the necessity.

@@

% \cleardoublepage
\subsection{\label{sec:linear_regression}Linear Regression}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\textwidth]{plots/regression/regression_G00010001.pdf}
    \caption{(a) Network geometry used for linear regression training. The network comprises $17$ edges connecting $10$ nodes, including one output node (purple dot) and three input nodes (green dots). Among the three input nodes, one is used to ground the circuit ($V_2 = 0V$), one ($V_4$) provides a constant voltage source for bias, and the remaining one ($V_4$) is used to input the training data point. (b) Test cost function evaluated every $10$ training steps for four different weight choices. Insets show visual representations of the network’s output trained with varying channel lengths at three different training steps.}\label{fig:lin_regr1}
\end{figure*}

@@

The performance of the algorithm on this task is shown in \cref{fig:lin_regr2}(b), where the learning rates for the different weights have been fine-tuned to prevent overshooting while still maintaining comparable convergence speed. From the behaviour of the cost function, it is evident that the choice of weight type has a decisive impact on the network’s ability to learn a complex task. Specifically, training the lengths of the channels on the edges results in successful convergence, whereas training the node pressures proves to be less efficient. For a visual interpretation, insets in \cref{fig:lin_regr1}(b) show the state of the training at three different time steps when the selected weight is the length of the channel.
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\textwidth]{plots/regression/regression_G00010002.pdf}
    \caption{(a) Network geometry used for linear regression training. It is equivalent to the geometry shown in \cref{fig:lin_regr1}, but the direction of the memristors highlighted in orange has been inverted. (b) Test cost function evaluated every $10$ training steps for four different weight choices. Insets show visual representations of the network’s output trained with varying node ion density at two different training steps.}\label{fig:lin_regr2}
\end{figure*}

To study the influence of memristor orientation on training performance, we retained the same network geometry but inverted the orientation of some memristors, as shown in \cref{fig:lin_regr2}(a). Specifically, if in the original configuration the base of a memristor pointed to one node and the tip to another, we reversed this direction. This modification significantly affects the behaviour of the network. As shown in \cref{fig:lin_regr2}(b), the new orientation alters the trainability of different weight types: in this configuration, training node concentration leads to better performance, while training edge length becomes less efficient. The underlying reason is that inverting the orientation of a channel also inverts the sign of the external stimulus applied to it, since the stimuli are defined node-wise and the difference is always taken as ''base minus tip".

This reversal changes both the shape and the conductivity range over which the function $g_{\infty}(\boldsymbol{\Delta})$ is defined, thereby limiting or enabling the conductivity values the channel can reach. A network with well-oriented channels can therefore access conductivity ranges that make the training solvable, while a poorly oriented network may be constrained to regions where solutions are not found. To exemplify this point, Figure 2 of Ref.\cite{Rene_PressureSensitive} shows $g_{\infty}$ as a function of pressure for positive and negative potential drops, assuming a zero concentration drop. The plot reveals two completely disjoint conductivity ranges depending on the sign of the potential drop, which explains why training on pressure alone may be challenging. This highlights the need for deeper exploration of network geometry when selecting the type of weight to train. In this regard, we observe that the edges whose orientation was inverted are directly connected to either the input or the output nodes. This suggests that particular attention should be given to these edges in future explorations, as they may have a bigger impact on trainability.

@@

We proposed a physical neural network composed of nodes connected by aqueous memristors. These devices exhibit conductance that responds to external stimuli of various physical natures, such as electrical, mechanical, and chemical. Additionally, their conductivity is sensitive to changes in the geometric parameters of their channel-like structure. Our motivation for using these devices lies in the ability to exploit these dependencies to control the conductance of each memristor in the network. This enables us to manipulate an input electrical signal in a programmable way, according to a specific computational objective.

@@

Training larger and more complex memristor networks is also feasible with the algorithm we proposed. We demonstrated that a network composed of 9 nodes and 15 edges can be successfully trained for more elaborated input-output voltage mapping using all the weight types considered.

We then challenged a more complex network, with 10 nodes and 17 edges, to perform a linear regression task with an offset. In this setting, we observed that the trainability of the network using a specific weight type strongly depends on the network's geometry. First, we showed that different geometries yield different final training accuracies when using the same weight type. Then, keeping the geometry fixed, we demonstrated that the orientation of the memristors also plays a crucial role: reversing the direction of just three memristors connected to either input or output nodes improved the training precision when using concentration as the trained weight by three orders of magnitude.

Given that the expressiveness of deep neural networks grows exponentially with depth\cite{big_neuralnet1, big_neuralnet2}, we interpret the limitations encountered in training linear regression tasks as a consequence of the small size of the network. Larger networks offer more degrees of freedom and can, in principle, enable broader exploration of the solution space. However, their experimental realization becomes complex. In this work, we showed that an appropriately chosen geometry adapted to a specific training objective can be sufficient to make even a small networks trainable. A deeper analysis of the role of geometry in determining trainability is still needed in the context of physical learning.
